---
title: "IHC Heterogeneity: How Much Does a Biopsy Tell You About the Tumour?"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{IHC Heterogeneity: How Much Does a Biopsy Tell You About the Tumour?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The Question

A pathologist scores Ki-67 on a core biopsy and the report carries that number forward as though
it described the tumour. It describes the core. If the marker is heterogeneously expressed, a
second core from the same tumour would have given a different answer, and the difference between
those two numbers is not measurement error — it is the tumour.

`ihcheterogeneity` quantifies that gap. Give it a reference measurement (whole section, or a
hot-spot score) and two or more regional measurements from the same cases, and it reports how well
the regions reproduce the reference, whether they do so with a systematic offset, and how much of
the total variance is sampling rather than real between-case difference.

This matters clinically wherever a threshold is applied to a single sample: Ki-67 cut-offs in
neuroendocrine tumours and breast cancer, PD-L1 scoring, HER2 quantification. A marker with high
within-tumour variability crossing a threshold on one core is not the same finding as one that
sits comfortably on one side of it.

## Data Layout

One row per case. One column for the reference measurement, one column per region.

```{r data}
set.seed(11)
n <- 40
ihc <- data.frame(CaseID = sprintf("C%02d", 1:n))
ihc$Whole <- round(pmin(100, pmax(0, rnorm(n, 45, 18))), 1)   # whole-section Ki-67 %

jitter_from <- function(mu, sd) round(pmin(100, pmax(0, mu + rnorm(n, 0, sd))), 1)
ihc$Biopsy1 <- jitter_from(ihc$Whole, 8)    # a well-sampled region
ihc$Biopsy2 <- jitter_from(ihc$Whole, 11)
ihc$Biopsy3 <- jitter_from(ihc$Whole, 14)   # a poorly representative one
ihc$Compartment <- factor(rep(c("Tumour centre", "Invasive front"), each = n / 2),
                          levels = c("Tumour centre", "Invasive front"))   # for the compartment example

head(ihc)
```

`wholesection` is optional. Without it the analysis compares the regions against each other, which
answers "do two cores agree?" rather than "does a core represent the tumour?" — a weaker question,
but the only one available when no whole-section score exists.

## A Complete Run

```{r analysis}
res <- ihcheterogeneity(
  data         = ihc,
  wholesection = "Whole",
  biopsy1      = "Biopsy1",
  biopsy2      = "Biopsy2",
  biopsy3      = "Biopsy3",
  # biopsy4 and `biopsies` take a fourth region and any number beyond that
  analysis_type            = "comprehensive",  # reproducibility | variability | comprehensive
  sampling_strategy        = "random",         # random | systematic | stratified | unknown
  cv_threshold             = 20,               # largest acceptable mean per-case CV (%)
  correlation_threshold    = 0.7,              # smallest acceptable Spearman correlation
  bias_margin              = 5,                # largest acceptable systematic difference (% of the reference mean)
  variance_components      = TRUE,
  sample_size_planning     = TRUE,
  generate_recommendations = TRUE,
  show_variability_plots   = TRUE
)
```

### Reproducibility

```
                                                  metric  value ci_lower ci_upper  interpretation
                     Mean Regional-Reference Correlation  0.760    0.559    0.877  Meets your threshold (>= 0.7)
              Spearman correlation: Biopsy1 vs reference  0.862    0.727    0.933  Meets your threshold (>= 0.7)
              Spearman correlation: Biopsy2 vs reference  0.757    0.554    0.875  Meets your threshold (>= 0.7)
              Spearman correlation: Biopsy3 vs reference  0.662    0.414    0.818  Below your threshold (< 0.7)
                           ICC(2,1) - absolute agreement  0.663    0.527    0.783  Moderate reliability
                     ICC(3,1) - consistency (bias-blind)  0.658    0.521    0.779  Ignores systematic offset
                         Mean Inter-Regional Correlation  0.554       NA       NA  Below your threshold (< 0.7)
 Mean Coefficient of Variation (%) - region vs reference 26.741       NA       NA  High variability
                                  Median per-case CV (%) 23.178       NA       NA  Robust to a few extreme cases
```

Each region is graded on its own. Biopsy3 (0.662) is below the 0.7 threshold as a point estimate,
but its 95% CI reaches 0.818, so it is not *shown* to be below it; only a region whose whole
interval lies below the threshold counts against the verdict.

Both ICCs are reported deliberately. **ICC(3,1)** asks whether the regions rank cases the same way;
**ICC(2,1)** asks whether they give the same number. A large gap between them means a systematic
offset — one region reads consistently high or low — which ranking-based agreement hides. Here they
are nearly identical (0.658 vs 0.663), so there is no meaningful offset; the disagreement is
scatter, not bias.

Note that mean inter-regional correlation (0.554) is well below the region-reference correlation
(0.760). Two cores agree with each other *less* than either agrees with the whole section, which is
what you would expect when each core samples a different part of a heterogeneous tumour.

### Sampling Bias

```
                       comparison  n mean_diff ci_lower ci_upper loa_lower loa_upper p_value
             Biopsy1 vs reference 40     0.287   -2.154    2.729   -14.675    15.250   0.813
             Biopsy2 vs reference 40     0.110   -3.316    3.536   -20.886    21.106   0.949
             Biopsy3 vs reference 40     1.400   -3.355    6.155   -27.741    30.541   0.555
 Mean of all regions vs reference 40     0.599   -1.395    2.594   -11.625    12.823   0.547

                       comparison                                               clinical_impact
             Biopsy1 vs reference   +0.7% (90% CI -4.5% to 5.9%): inconclusive at the 5% margin
             Biopsy2 vs reference   +0.3% (90% CI -7.0% to 7.6%): inconclusive at the 5% margin
             Biopsy3 vs reference  +3.6% (90% CI -6.5% to 13.7%): inconclusive at the 5% margin
 Mean of all regions vs reference   +1.5% (90% CI -2.7% to 5.8%): inconclusive at the 5% margin
```

`clinical_impact` is the column to read, not `p_value`. Each difference is judged against
`bias_margin`, here 5% of the reference mean: it is ruled out when its 90% CI lies inside the
margin (two one-sided tests), shown to be material when a Bonferroni-adjusted CI lies entirely
beyond it, and inconclusive otherwise. All four rows are inconclusive. The estimates are small
(0.3% to 3.6%), but with 40 cases the 90% CIs still reach past 5% (Biopsy3 up to 13.7%), so a
non-significant p-value here does not show that the regions are unbiased.

The limits of agreement say how far a single case can differ: for Biopsy3 between -27.7 and
+30.5 points. The table also gives the 95% CI of each limit (Bland & Altman 1999; for Biopsy3,
-35.9 to -19.5 and 22.3 to 38.7), which is wide with 40 cases.

### Variance Components

```
                       component variance percentage                  contribution
           Between-Case Variance    196.65       65.8             Major contributor
 Within-Case Variance (Sampling)    102.19       34.2     High sampling variability
                 Method Variance      0.00        0.0 Small share of total variance
                  Total Variance    298.84      100.0 Sum of all variance components
```

This is the number to quote in a validation report. **34.2% of the total variance is within-case**
— that is the proportion of the spread in your dataset that comes from *where you sampled*, not
from which patient you sampled. A cut-off applied to a single core inherits all of it.

### Sample Size for ICC Precision

```
                                                 scenario planning_icc width_current n_w20 n_w10
      Planning ICC 0.75 (lower limit of good reliability)        0.750         0.208    44   171
 Planning ICC 0.90 (lower limit of excellent reliability)        0.900         0.095    10    37
                           Observed ICC(3,1), consistency        0.658         0.261    68   266
```

This is the planning question an agreement study has: how many cases pin the ICC down. With four
measurements per case (the reference and three regions), 44 cases give an expected 95% CI width of
0.20 if the true ICC is 0.75; at the ICC observed here (0.66) the same precision needs 68 cases,
and a width of 0.10 needs 266. The widths are expected values (Bonett 2002), so about half of the
studies of that size obtain a wider interval.

### The Verdict

The Clinical Assessment reads **MODERATE SAMPLING**: the mean correlation meets 0.7, but the mean
per-case CV (26.7%) is above the 20% threshold and within only the relaxed band of 30%
(1.5 x the threshold). That relaxed band is a heuristic of this analysis, not a published
criterion; the assessment says so, and it suggests averaging more than one region per case.

## Spatial Compartments

When regions are labelled by anatomical compartment — invasive front versus tumour centre, say —
`spatial_id` names that column and the two compartment options compare them:

```{r compartments}
res_spatial <- ihcheterogeneity(
  data                = ihc,
  wholesection        = "Whole",
  biopsy1             = "Biopsy1",
  biopsy2             = "Biopsy2",
  spatial_id          = "Compartment",
  compareCompartments = TRUE,
  compartmentTests    = TRUE
)
```

```
                                   test_type statistic df1 df2 p_value
           Kruskal-Wallis test (per-case CV)     0.354   1  NA   0.552
 Brown-Forsythe test (spread of per-case CV)     0.125   1  38   0.725
        Kruskal-Wallis test (per-case means)     0.089   1  NA   0.766
```

The compartments in this example were assigned by row, not by biology, so no test detects a
difference, as it should. The Kruskal-Wallis test on the per-case CV is the one that asks whether
one compartment is more heterogeneous than another; the other two describe the spread of the CVs
and the biomarker level.

A significant compartment difference is a different finding from general heterogeneity: it says the
marker varies *systematically* by location, so where the biopsy is taken from is a protocol
decision, not a matter of chance.

## Reading the Result

- `analysis_type` sets how much is shown: `reproducibility` for the correlations, ICCs, CV and
  systematic differences; `variability` adds the variance components and the plots;
  `comprehensive` (the default) adds sample-size planning as well.
- `bias_margin` (default 5, percent of the comparison mean) is the largest systematic difference
  you would accept. Choose it, like the two thresholds below, before looking at the results.
- `cv_threshold` (default 20) and `correlation_threshold` (default 0.8) set the bars against which
  results are labelled acceptable. They are conventions, not standards — if your assay has a
  published reproducibility requirement, use that number instead.
- `sampling_strategy` documents how the regions were chosen. It does not change the arithmetic, but
  it belongs in the record: results from systematically placed cores do not generalise to randomly
  placed ones.
- `showSummary` adds a plain-language summary and `showGlossary` a statistics glossary, both useful
  when the output is going to a tumour board rather than a statistician.

## Limitations

The analysis describes the cases you measured. It cannot tell you that two cores are sufficient in
general — only that, in this cohort, 34% of the variance lay within cases rather than between them. Extending that
to a clinical protocol needs a prospectively defined validation study with the threshold and the
decision it drives specified in advance.
