---
title: "IHC Heterogeneity: How Much Does a Biopsy Tell You About the Tumour?"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{IHC Heterogeneity: How Much Does a Biopsy Tell You About the Tumour?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The Question

A pathologist scores Ki-67 on a core biopsy and the report carries that number forward as though
it described the tumour. It describes the core. If the marker is heterogeneously expressed, a
second core from the same tumour would have given a different answer, and the difference between
those two numbers is not measurement error — it is the tumour.

`ihcheterogeneity` quantifies that gap. Give it a reference measurement (whole section, or a
hot-spot score) and two or more regional measurements from the same cases, and it reports how well
the regions reproduce the reference, whether they do so with a systematic offset, and how much of
the total variance is sampling rather than real between-case difference.

This matters clinically wherever a threshold is applied to a single sample: Ki-67 cut-offs in
neuroendocrine tumours and breast cancer, PD-L1 scoring, HER2 quantification. A marker with high
within-tumour variability crossing a threshold on one core is not the same finding as one that
sits comfortably on one side of it.

## Data Layout

One row per case. One column for the reference measurement, one column per region.

```{r data}
set.seed(11)
n <- 40
ihc <- data.frame(CaseID = sprintf("C%02d", 1:n))
ihc$Whole <- round(pmin(100, pmax(0, rnorm(n, 45, 18))), 1)   # whole-section Ki-67 %

jitter_from <- function(mu, sd) round(pmin(100, pmax(0, mu + rnorm(n, 0, sd))), 1)
ihc$Biopsy1 <- jitter_from(ihc$Whole, 8)    # a well-sampled region
ihc$Biopsy2 <- jitter_from(ihc$Whole, 11)
ihc$Biopsy3 <- jitter_from(ihc$Whole, 14)   # a poorly representative one

head(ihc)
```

`wholesection` is optional. Without it the analysis compares the regions against each other, which
answers "do two cores agree?" rather than "does a core represent the tumour?" — a weaker question,
but the only one available when no whole-section score exists.

## A Complete Run

```{r analysis}
res <- ihcheterogeneity(
  data         = ihc,
  wholesection = "Whole",
  biopsy1      = "Biopsy1",
  biopsy2      = "Biopsy2",
  biopsy3      = "Biopsy3",
  # biopsy4 and `biopsies` take a fourth region and any number beyond that
  analysis_type            = "comprehensive",  # reproducibility | bias | variability | comprehensive
  sampling_strategy        = "random",         # random | systematic | stratified | unknown
  cv_threshold             = 20,               # CV% above this is flagged as unacceptable
  correlation_threshold    = 0.7,              # minimum acceptable region-reference correlation
  variance_components      = TRUE,
  power_analysis           = TRUE,
  generate_recommendations = TRUE,
  show_variability_plots   = TRUE
)
```

### Reproducibility

```
                             metric     value ci_lower ci_upper  interpretation
 Mean Regional-Reference Correlation     0.760       NA       NA  Good representativeness
       ICC(2,1) - absolute agreement     0.663    0.527    0.783  Moderate reliability
 ICC(3,1) - consistency (bias-blind)     0.658    0.521    0.779  Ignores systematic offset
     Mean Inter-Regional Correlation     0.554       NA       NA  Variable sampling
```

Both ICCs are reported deliberately. **ICC(3,1)** asks whether the regions rank cases the same way;
**ICC(2,1)** asks whether they give the same number. A large gap between them means a systematic
offset — one region reads consistently high or low — which ranking-based agreement hides. Here they
are nearly identical (0.658 vs 0.663), so there is no meaningful offset; the disagreement is
scatter, not bias.

Note that mean inter-regional correlation (0.554) is well below the region-reference correlation
(0.760). Two cores agree with each other *less* than either agrees with the whole section, which is
what you would expect when each core samples a different part of a heterogeneous tumour.

### Sampling Bias

```
                   comparison mean_diff p_value effect_size clinical_impact
        Region 1 vs Reference     0.288   0.813      0.0369   Minimal (<5%)
        Region 2 vs Reference     0.110   0.949      0.0101   Minimal (<5%)
        Region 3 vs Reference     1.400   0.555      0.0923   Minimal (<5%)
 Mean of Regions vs Reference     0.599   0.547      0.0942   Minimal (<5%)
```

`clinical_impact` is the column to read, not `p_value`. A non-significant p-value on 40 cases does
not establish that the bias is small; the effect size and the stated impact band do.

### Variance Components

```
                       component variance percentage                  contribution
           Between-Case Variance    196.65       65.8             Major contributor
 Within-Case Variance (Sampling)    102.19       34.2     High sampling variability
                 Method Variance      0.00        0.0 Negligible method differences
                  Total Variance    298.84      100.0
```

This is the number to quote in a validation report. **34.2% of the total variance is within-case**
— that is the proportion of the spread in your dataset that comes from *where you sampled*, not
from which patient you sampled. A cut-off applied to a single core inherits all of it.

### Power

```
              scenario effect_size power required_n                     recommendation
  Small Effect (r=0.1)        0.10 0.094        783    Substantial sample increase
 Medium Effect (r=0.3)        0.30 0.469         85    Substantial sample increase
  Large Effect (r=0.5)        0.50 0.916         30       Adequate power achieved
  Observed Effect Size        0.91 1.000          7  Post-hoc (observed) power
```

The last row is labelled as post-hoc for a reason: power computed from the observed effect is a
restatement of the p-value, not evidence that the study was adequately powered. Use the first three
rows for planning.

## Spatial Compartments

When regions are labelled by anatomical compartment — invasive front versus tumour centre, say —
`spatial_id` names that column and the two compartment options compare them:

```{r compartments}
res_spatial <- ihcheterogeneity(
  data                = ihc_with_regions,
  wholesection        = "Whole",
  biopsy1             = "Region_A",
  biopsy2             = "Region_B",
  spatial_id          = "Compartment",
  compareCompartments = TRUE,
  compartmentTests    = TRUE
)
```

A significant compartment difference is a different finding from general heterogeneity: it says the
marker varies *systematically* by location, so where the biopsy is taken from is a protocol
decision, not a matter of chance.

## Reading the Result

- `analysis_type` narrows the output when you only need part of it: `reproducibility` for
  correlations and ICCs, `bias` for the systematic-offset tests, `variability` for CV and variance
  components, `comprehensive` (the default) for all of them.
- `cv_threshold` (default 20) and `correlation_threshold` (default 0.8) set the bars against which
  results are labelled acceptable. They are conventions, not standards — if your assay has a
  published reproducibility requirement, use that number instead.
- `sampling_strategy` documents how the regions were chosen. It does not change the arithmetic, but
  it belongs in the record: results from systematically placed cores do not generalise to randomly
  placed ones.
- `showSummary` adds a plain-language summary and `showGlossary` a statistics glossary, both useful
  when the output is going to a tumour board rather than a statistician.

## Limitations

The analysis describes the cases you measured. It cannot tell you that two cores are sufficient in
general — only that, in this cohort, two cores left 34% of the variance unexplained. Extending that
to a clinical protocol needs a prospectively defined validation study with the threshold and the
decision it drives specified in advance.
